---
title: "rtmbGMACS CTL: initial abundance specification"
author:
- name: William T. Stockhausen
  affiliations:
    - id: afsc-refm
      name: "Alaska Fisheries Science Center"
      department: "Resource Ecology and Fisheries Management"
      address: 
        - "7600 Sand Point Way N.E."
        - "Seattle, Washington 98115-6349"
date: '`r format(Sys.time(), "%b %e, %Y")`'
fontsize: 11pt 
number-sections: true
number-depth: 2
reference-location: document
bibliography: '`r path.expand("~/Work/Projects/Bibliography/AllRefs.bib")`'
csl: '`r system.file("files/CJFAS.csl",package="wtsQMD")`'
link-citations: true
crossref:
  chapters: false      # prepend label reference numbers by chater number?
  fig-title: Figure    # for caption: default is "Figure")
  tbl-title: Table     # for caption: default is "Table")
  title-delim: "."     # for caption: default is ":")
  fig-prefix: Figure   # for in-text (use [-@fig-ref] to drop prefix in text)
  tbl-prefix: Table    # for in-text (use [-@tbl-ref] to drop prefix in text)
  fig-labels: arabic    # options are arabic, roman, roman i, alpha x, alpha X; latter two starting from x/X
  tbl-labels: arabic    # options are arabic, roman, roman i, alpha x, alpha X; latter two starting from x/X
  subref-labels: alpha a # options are arabic, roman, roman i, alpha x, alpha X; latter two starting from x/X
  ref-hyperlink: true    # references are hyperlinked if true
format: 
  html: 
    df-print: paged
    toc: true
    toc-location: right
    fig-width: 8
    fig-asp: 0.4
    fig-dpi: 100
    embed-resources: true
    include-in-header: 
      - '`r system.file("files/in-line_math_mode.html",package="wtsQMD")`'
  pdf:
    documentclass: scrartcl
    toc: false
    fig_crop: false
    keep-tex: false
    geometry:
      - left=1.0in
      - right=1.0in
      - top=1.0in
      - bottom=1.0in
      - textwidth=6.5in
      - showframe=false
    include-in-header: 
       #include '\usepackage{nopageno}' below in header to turn off page numbering
      - text: |
          \usepackage{placeins}
          \extrafloats{500}
          \maxdeadcycles=10000
          \usepackage{fontspec}
          \usepackage{multicol}
          \usepackage{hhline}
          \newlength\Oldarrayrulewidth
          \newlength\Oldtabcolsep
          \usepackage{nopageno}
      - file: '`r system.file("files/ltx_ExtraLatexIncludes.tex",package="wtsQMD")`'
echo: false
warning: false
results: 'hide'
keep-md: true
keep-yaml: false
editor: source
editor_options: 
  chunk_output_type: console
concordance: true
params:
  setup: !expr 'system.file("files/qmd_setup.R",package="wtsQMD")'
  testing: false
  reorderTables: false   #--issue with finding "range" of table values  (don't use)
  reorderFigures: false  #--issue with finding "range" of figure values (don't use)
---
<!-- IMPORTANT: if used as a child doc, all chunk labels must be unique within the overall document -->

<!-- 
  NOTEs: 
     * child_path$peek() gives path to current script.
     * before starting a child document, do 
         "child_path$push(file.path(child_path$peek(),rel_path))" first, 
         where rel_path is the relative path to the child
     * after a child document has finished, do "child_path$pop()" to return to current path value
-->

<!-- if not child doc, set up required objects -->
```{r}
#| label: CTL-InitialAbundance_setup
#| eval: !expr '!knitr::opts_knit$get("child")'
#| results: asis
  testing = params$testing;
  if (testing) cat(params$setup,"\n\n")
  source(params$setup);
  if (testing) cat("root = ",root,"\n\n")
  reorderTables  = params$reorderTables;
  reorderFigures = params$reorderFigures;
```

```{r}
#| label: setup_CTL-InitialAbundance
#| results: asis
  require(rtmbGMACS);
  require(ggplot2);
  require(kableExtra);
  require(tables);
  thm = wtsPlots::getStdTheme();
  options("readr.show_col_types"=FALSE);
  if (testing) cat("dirThs =",child_path$peek(),"\n\n")
  if (!exists("tblno")) tblno = 0;#--index in LETTERS for tables in text
```

```{r}
#| label: CTL-InitialAbundance-ModelDimensions
#| eval: !expr '!knitr::opts_knit$get("child")'
vRs="EBS";
miz = as.character(seq(25,75,5));
mmz = as.character(seq(50,75,5));
fiz = as.character(seq(25,65,5));
fmz = as.character(seq(30,65,5));
vXs=list(male=list(imm=list(`new`=miz,
                             `old`=miz),
                    mat=list(`new`=mmz,
                             `old`=mmz,
                             `very`=mmz)
                  ),
          female=list(imm=list(`new`=fiz,
                               `old`=fiz),
                      mat=list(`new`=fmz,
                               `old`=fmz)
                     )
         ); attr(vXs,"dmnms")<-c("x","m","p","z");
ys = as.character(2015:2024); names(ys) = ys;
ss = as.character(1:2);       names(ss) = ss;
dmsC    = createSparseDimsMap(r=vRs,x=vXs);            #--DimsMap for stock categories
dmsYS   = createSparseDimsMap(y=ys,s=ss);              #--DimsMap for years and seasons
dmsYSC  = createSparseDimsMap(y=ys,s=ss,r=vRs,x=vXs);  #--DimsMap for combination 
dimsZBC = dmsC |> dplyr::mutate(lft=as.numeric(z)-2.5,mid=as.numeric(z),rgt=as.numeric(z)+2.5);
fleets = c("Directed","Bycatch","Survey"); #--fishery and survey fleets
```

### Initial abundance {#sec-InitialAbundance}

The following options are available to define initial abundance in the model:

  - "zero_pop" - no parameters
  - "steady-state_unfished"
  - "steady-state_fished"
  - "free_parameters_v1" - freely estimated parameters
  - "free_parameters_v2" - freely estimated parameters
  - "fixed_abundance"    - fixed values
  
```{r}
#| label: CTL-InitialAbundanceZeroPop
#| echo: true
str<-paste0('
  #--rtmbGMACS initial N specifications----
  PROCESS  INITIAL_N
  #--calculation options
  ##--"zero_pop" - no parameters
  ##--"steady-state_unfished"
  ##--"steady-state_fished"
  ##--"free_parameters_v1" - freely estimated parameters
  ##--"free_parameters_v2" - freely estimated parameters
  ##--"table"              - table of values
  zero_pop #--selected calculation option
  
  ##--zero_pop: no further specification needed
');
```

```{r}
#| label: CTL-InitialAbundanceSteadyStateUnfished
#| echo: true
str<-paste0('
  #--rtmbGMACS initial N specifications----
  PROCESS INITIAL_N
  steady-state_unfished #--selected calculation option
');

```

```{r}
#| label: CTL-InitialAbundanceSteadyStateFished
#| echo: true
str<-paste0('
  #--rtmbGMACS initial N specifications----
  PROCESS INITIAL_N
  steady-state_fished #--selected calculation option
  END_PROCESS
');

```

```{r}
#| label: CTL-InitialAbundanceFreeParamsV1
#| echo: true
str<-paste0('
  #--rtmbGMACS initial N specifications----
  PROCESS INITIAL_N
  free_parameters_v1 #--selected calculation option
  END_PROCESS
');

```

```{r}
#| label: CTL-InitialAbundanceFreeParamsV2
#| echo: true
str<-paste0('
  #--rtmbGMACS initial N specifications----
  PROCESS INITIAL_N
  free_parameters_v2 #--selected calculation option
  END_PROCESS
');

```

```{r}
#| label: CTL-InitialAbundanceFixedValues
#| echo: true
str<-paste0('
  #--rtmbGMACS initial N specifications----
  PROCESS INITIAL_N
  table      #--selected calculation option
  arithmetic #--scale ("arithmetic","log-scale","relative")
  millions   #--units
  scale_parameter
  IV  phase LB  UB  prior_fcn  p1  p2  
  
  r    x    m     p     z     IV   phase   LB     UB     prior_fcn     p1     p2
  END_PROCESS
');

```

### Allometry (weight-at-size)
### Recruitment
### Natural mortality
### Size/stage transitions 
#### Molt probability
#### Growth 
#### Maturation 
#### Transition matrices
### Selectivity and retention
### Fishing mortality 
### Survey catchability 
### Movement (TBD?)

```{r}
#| label: refs_CTL-InitialAbundance
#| eval: !expr '!knitr::opts_knit$get("child")'
#| results: asis
  cat("# References {-}\n")
  cat("::: {#refs}\n")
  cat(":::\n\n")
```

<!-- tables, if not child doc and lstTbls is not empty -->
```{r}
#| label: tables_CTL-InitialAbundance
#| eval: !expr '!knitr::opts_knit$get("child")'
#| results: asis
  source(system.file("files/printTablesSectionFromList.R",package="wtsQMD"));
```

<!-- figures, if not child doc and lstFigs is not empty -->
```{r}
#| label: figures_CTL-InitialAbundance
#| eval: !expr '!knitr::opts_knit$get("child")'
#| results: asis
  source(system.file("files/printFiguresSectionFromList.R",package="wtsQMD"));
```

